Computational methods in landscape ecology

### 2024-05-28 Sections/spatial-patterns.qmd
Adding some notes to spatial-patterns

Landscape ecology is fun [@Turner1989].
There are many things to explore [@With2019].

@@

@@

spatial patterns definitions in the literature (general)
?existance of spatial pattterns in many contexts, from geography, through biology and urban planning to medical studies
?relation of spatial patterns to spatial processes

spatial patterns in data (computational approach); its limitations and considerations
raster data
vector data: point, line, polygon
point clouds 
time series data

one variable vs. multivariable spatial patterns
categorical vs. continuous spatial patterns

resolution, scale, and extent in spatial patterns

short intro to the subsequent sections

!to conclusions that relation of spatial patterns to spatial processes is a golden grail of landscape ecology
!what with fractals?
!ai and deep learning still seems to be underused in landscape ecology

@@

## Surface metrics
<!-- Bosch, Hesselbarth -->

## Entropy
<!-- Nowosad, Bosch -->

intro to entropy
various concepts:

>>>>

@@

[Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022)
[Where are we now, and where are we going?](https://doi.org/10.1007/s10980-019-00889-6)

@@

- Spatial patterns definitions in the literature (general; Dale 2000)
- Spatial patterns in many contexts (e.g. geography, biology, urban planning, medical studies) <!-- MH: I would at least cut the medical bit -->
- Spatial patterns in landscape ecology
  - Resolution, scale, and extent in spatial patterns
- Relation of spatial patterns and spatial (ecological?) processes
- Relation of spatial patterns to spatial processes is a golden grail of landscape ecology

- Spatial patterns in data (computational approach)
  - Limitations and considerations
- Data types <!-- MH: Could be moved to the individual sub-sections? -->
  - Raster data
  - Vector data: point, line, polygon
  - Point clouds 
  - Time series data
  - One variable vs. multivariable spatial patterns
  - Categorical vs. continuous spatial patterns

!what with fractals?
!ai and deep learning still seems to be underused in landscape ecology

@@

- Traditional approach to describe patterns
- Quantify categorical raster data (patch mosaic model)
- Quantify composition and configuration
- Widely applied since FRAGSTATS release
- Several alternative tools available and still widely used

## Surface metrics
<!-- Bosch, Hesselbarth -->
[Surface metrics])(https://doi.org/10.1007/s10980-018-0685-1)

- More recent development
- Quantify continuous raster data (gradient surface models)
- More realistic representation of landscapes
- Based on molecular physics/medicine
- Not widely adopted yet (due to missing tools?)

## Entropy
<!-- Nowosad, Bosch -->
intro to entropy
various concepts:
